Parallelization of Push-based System for Molecular Simulation Data Analysis with GPU

### 2016-02-08 paper/paperSource.tex
automatic triggered changes

The proposed design has a pretty straightforward structure which is going to be explained in this section. Shortly, the system starts with original input files (or data) and gets consumed in a sequential manner by simple disk IO framework into random access memory by chunks, then, chunk gets uploaded to GPU device and processed, at some point, the results that need to be shown to the user are transported from GPU (where they initially get accumulated) to the main memory, and then might get written as well to some other output files.

\begin{figure}
 \centerline{ \includegraphics[width=0.65\columnwidth]{images/memory-organization.eps} }
 \caption{Data flow}
 \label{fg:data-flow}
\end{figure}

\subsection{Data flow}
Since the size of the data in the problem is huge, organization of the data flow is crucial in our case. As mentioned before, original data is retrieved from simulation software output files and in our case it is Gromacs Software's formatted trajectory files. Tool for decompressing those files to a flat format has been developed in the previous work\cite{mainPaper}.

\emph{\textbf{Data read - transform.}} As mentioned above, in order
for our system to be able to read the MS generated data, the data
needs to be transformed. The reason for this is following: the MS
data is stored in multiple files and possibly in different formats
as well. One such file holds the global data (identifying the system
and the simulation). Another file holds each frame's data. The frame
data contain general information about the frame, but the main part
is a sequential list of each atom's info, including atom's mass,
position, charge, number of electrons, velocities, forces, etc.
Another file (topology file) holds the molecule/residue info,
essentially identifying what atom belongs to which molecule.

So, in order to extract the data from these files, we have created a
sort of "extractor/transformer" of the attributes needed for the
execution of the queries in our system. This transformer,
essentially, is a separate piece of code that does three things: 1)
it reads the MS generated data that the MS system stores in one of
the often used MS file format (e.g., GROMACS); 2) it translates the
data into a format that our system can read (taking only the
information our system needs); and 3) it stores the data in a file
that has basic structure to it. So, in the end, the data transformer
produces a data file that our system takes as input. This code
serves as a connection between an MS system (e.g., GROMACS) and our
system. With this, our system can essentially be used as an add-on
to GROMACS or other simulation systems and help improve the
efficiency of the data analysis.

\textbf{\emph{Data organization in main memory.}}\footnote{This
paragraph talks only about the data organization used by one-body
queries. For two-body queries (e.g., SDH), the data organization is
discussed later on.} Once the data is in a format our system can
read, the data is being loaded into the main memory one frame at a
time. The organization of the in-memory particle's data is in the
form of a simple, two dimensional array where a single row
represents an atom in the system with all its attributes (e.g.,
coordinates, mass, charge, residue info, etc.). We also keep (in a
one dimensional array) crucial system's information for each frame,
like temperature, energy, pressure, etc. We have used such
structures because they are very suitable for simulating a
push-based type of system: a simple sequential read of the array
gives that on-line type of data stream. So, as the system reads the
array, it pushes the data onto the query-modules. As mentioned
earlier, the one-body (algebraic) queries will produce a final
result at the end of the first sequential read. However, the
two-body (holistic) functions, like SDH and RDF, cannot do this in a
single read of the data.\footnote{However, we have previously
deigned and created a data structure and an algorithm that can take
the advantage of a single data read and produce final results for
SDH computation~\cite{ytu:icde09,EDBT12}. We have incorporate this
into our system presented in this paper.} Therefore, each frame's
data array can be continuously read (in a loop manner) as many times
as a query needs it.
\subsection{Query modules}

As mentioned earlier, there are two types of functions/queries used
for analysis of MS systems: algebraic or one-body, and holistic or
two-body queries (these in general can be multi-body, but in this
paper we only deal with a two-body functions).

>>>>

The proposed design has a pretty straightforward structure which is going to be explained in this section. Shortly, the system starts with original input files (or data) and gets consumed in a sequential manner by simple disk IO framework into random access memory by chunks, then, chunk gets uploaded to GPU device and processed. At some point, the results that need to be shown to the user are transported from global memory of GPU (where they initially get accumulated) to the main memory, and then might get written as well to an output file. General data flow map is represented in Figure  \ref{fg:data-flow}.

\begin{figure}
 \centerline{ \includegraphics[width=0.65\columnwidth]{images/memory-organization.eps} }
 \caption{Data flow in the improved design of push-based system for MS analysis}
 \label{fg:data-flow}
\end{figure}

\subsection{Data input}
Since the size of the data in the problem is huge, organization of the data flow is crucial. As mentioned before, original data is retrieved from simulation software output files and in our case it is Gromacs Software's trajectory files\footnote{The file types, their structure, and their decompression has been well explained in the previous paper in section "Data orgranization in main memory"}. For a simulation, there can be several files of different format and usually they are also compressed without any loss of information. Tool for decompressing those files to a flat format for each frame and atom has been developed in the previous work\cite{mainPaper}.

In order to experiment with different input cases, the input files have been generated with a tool that we have built in Python language. Thus, we are able to generate mock data files with any number of frames and atoms, and the values are random and can be adjusted with needs of the experiment (for example, for cases when we want to analyze bucket widths in a limited space, the coordinates are generated in range of $0 - 1$, etc.). As it has already been mentioned earlier, we don't expect the data to be distributed with a certain behavior, the importance of having this kind of tool to generate mock data is having a variety of random data sets that could reveal any issues related to values provided to the program, on top of that, having the actual file written and then being read has been a part of the experiments to have more realistic use case. In the future, the same script for generating the mock data could be used for regression and performance tests in a distributed implementation of this design as a single source.

\subsection{Data flow architecture}

MS data roughly can be represented as a list of atoms with physical properties aggregated as frames. Thus, we consider a single frame as a chunk which will be loaded into the memory\footnote{Memory in this design might be interpreted differently, since we have a separate device memory on GPUs, but it will be explained better later} and pushed through our network of query. In this case, it is very convenient, because there are almost no dependencies between frames, rather than list of atoms in the same frame, which potentially again could create additional disk I/O overhead.

Having defined the chunk, we load it from flat format file to RAM. A single atom data structure has all the attributes such as charge, mass, velocity, type of atom, coordinates, and so on. In other words, we have an array of atoms, and most of the attributes (in case of data structure, it is a class field), are simply primitive floats or doubles depending on precision requirements. Next, the same chunk of data with the same structure is loaded onto GPU global memory. GPU devices, in general, have about 4-5 types of memory, and global memory is the biggest and slowest. Consequently, depending on algorithm or query we load parts of data chunk to another types of memory called shared memory. It depends on the actual query, because the same part of chunk can be loaded and unloaded in regards with shared memory several times, especially with SDH.

\subsection{Query modules}
As mentioned earlier, there are two types of functions/queries used
for analysis of MS systems: algebraic or one-body, and holistic or
two-body queries (these in general can be multi-body, but in this
paper we only deal with a two-body functions).
